Introductory statistics with `intRo`

### 2016-05-26 jcgs-paper-2016/response.Rmd

---
title: "Response to reviewers"
author: "Eric Hare, Andee Kaplan"
date: "May 26, 2016"
output: html_document
---

# Reviewer 1

This is an intelligent paper describing interesting software, but it (and the associated software) suffer from three problems:

(1) The paper seems to me inappropriate for the JCGS. Because it focuses purely on software, I think that (once the other issues are addressed) it would be more appropriate for the Journal of Statistical Software or the R Journal.

(2) Although the software looks promising, unless I've missed something, it currently has very limited functionality -- indeed, not enough to sustain most introductory statistics courses. This isn't a permanent limitation, however, since the authors have created interesting infrastructure for extending the software. In its current state, the software doesn't quite look ready for prime time.

(3) The authors ignore other (and at present, much more capable) GUIs for R -- the R Commander and Deducer. Since these are the most direct competitors to intRo -- more so than, e.g., Excel and JMP -- it seems obvious that the authors should address similarities, differences, advantages, and disadvantages. One obvious difference is that these other GUIs run on individual computers rather than being web-based, and it would also make sense to address more generally the advantages and disadvantages of the two approaches. (For example, in my tests of intRo, I noticed slow response, even with very small data sets. Of course, it's hard to know to what to attribute that -- it could as easily have been a network issue as slowness at the server end.)

# Reviewer 2

The “Article Assessment” questions on the report form don’t work for a paper like this.
The authors report on some very impressive work. The simplicity of conception and execution is one of its biggest strengths. Their motivations are compelling, as is their system design and execution but I think the story-telling can be improved considerably.

I think the authors need to decide, “What sort of paper am I writing?” Included in the manuscript are elements of “a paper about teaching introductory statistics using a software system”, elements of “a paper about designing and building a simple extensible system”, and “a paper about how to use and extend a particular system”.

I think the most compelling story for the Journal of Computational and Graphical Statistics is the designing and building of a simple online system for intro stat that is easy to learn, use and extend – and teaches R as a side effect. This would leave room for a sister publication in something like the Journal of Statistics Education which actually focusses on teaching aspects.

If you buy this conception, the teaching-related aspects should be shortened, shaved of much detail and form the basis of an introductory motivational section that leads into a design brief and a plan of attack (using shiny etc). There needs to be high-level discussion of how the system is conceived and put together accompanied by high-level diagrams before descending into code fragments (much detailed code-fragment discussion could go to appendices). The paper just needs to convey the sense of how things are done. Actual details of how to customise and extend the system are provided in the system’s documentation.

Detailed comments follow. (Perhaps unfortunately, many draw on what I have said above.):
Everywhere – distinguish between user=student (who uses what is served up) and user=teacher (who customises what is served up)

p. 2: Another online analysis system implemented in shiny is iNZight Lite at http://lite.docker.stat.auckland.ac.nz/. It is more ambitious and therefore more complex. Also it does not have the “exposing the R code” capability.

p 3 bottom: Remove parentheses around “(and its corresponding code)” because this is actually one of the key features. R Commander should enter the discussion around here (even though it isn’t online)

P 5: Fig 2 shows a simple design that should be easy for casual users (as intro students really are) to get their heads around. Not all that keen on the use of “statistics” as a category heading in Panel 2 because it reinforces the ghettoizing of the word conflicting with a message that says everything you are doing here is statistics.

P 6 top: When you look at the actual R code generated it can be quite complex with its pipe structures etc which I would have thought does rather limit its introductory value.

P 6-7: discussion of particular modules should be in a “for example” vein. There is no need to justify what particular features are and are not included. (That would be in a teaching paper.)

P 7 top: Being able to choose only plot types that are “compatible” with the variables used does, however, preclude learning experiences of the form “this thing you want to do is not appropriate for this type of data.”

P 8: I think this descends to details too quickly. The big features are designing for extensibility, with conceptions about switching modules on and off and adding modules, and the interface as presented being assembled on the fly when the system starts up.

P 9-10 looks like appendix stuff to me.

Section 3 discusses downloading onto a local machine before customising the hosted online system which should be the focus (again conception and issues involved before any details).

P 11 near bottom: “as an application on the user’s account” puzzled me -- needs to say shinyapps.io account
Section 3.2 mainly belongs in up-front motivation. Table 1 is an implementation mechanism.
P 14 top: “currently unsupported” justifications are more relevant to a teaching paper than a software design paper.
P 14 Last para Section 3: could go into motivational discussion.
Section 4.1 is a big part of what I see as the real story but descends to details too quickly.
Most of Section 4.2 looks to me like more motivation, except the interpolate stuff which is an implementation mechanism
Section 4.3: more “main story”.

### 2016-05-26 jcgs-paper-2016/response.Rmd

(3) The authors ignore other (and at present, much more capable) GUIs for R -- the R Commander and Deducer. Since these are the most direct competitors to intRo -- more so than, e.g., Excel and JMP -- it seems obvious that the authors should address similarities, differences, advantages, and disadvantages. One obvious difference is that these other GUIs run on individual computers rather than being web-based, and it would also make sense to address more generally the advantages and disadvantages of the two approaches. (For example, in my tests of intRo, I noticed slow response, even with very small data sets. Of course, it's hard to know to what to attribute that -- it could as easily have been a network issue as slowness at the server end.)

# Reviewer 2

The “Article Assessment” questions on the report form don’t work for a paper like this.
The authors report on some very impressive work. The simplicity of conception and execution is one of its biggest strengths. Their motivations are compelling, as is their system design and execution but I think the story-telling can be improved considerably.

**We appreciate the very kind comments and hope that this revision has improved the story being told.**

I think the authors need to decide, “What sort of paper am I writing?” Included in the manuscript are elements of “a paper about teaching introductory statistics using a software system”, elements of “a paper about designing and building a simple extensible system”, and “a paper about how to use and extend a particular system”. I think the most compelling story for the Journal of Computational and Graphical Statistics is the designing and building of a simple online system for intro stat that is easy to learn, use and extend – and teaches R as a side effect. This would leave room for a sister publication in something like the Journal of Statistics Education which actually focusses on teaching aspects.

**We have taken care to significantly shift the focus away from the teaching aspects, and towards the extensible software system aspects. In particular, we have removed the lengthy teaching discussion and moved elements of it to the motivational section. We have also trimmed much of the content involving supported and unsupported functionality. Specific changes are noted in the responses to the further points raised below.**

@@

If you buy this conception, the teaching-related aspects should be shortened, shaved of much detail and form the basis of an introductory motivational section that leads into a design brief and a plan of attack (using shiny etc). There needs to be high-level discussion of how the system is conceived and put together accompanied by high-level diagrams before descending into code fragments (much detailed code-fragment discussion could go to appendices). The paper just needs to convey the sense of how things are done. Actual details of how to customise and extend the system are provided in the system’s documentation.

Everywhere – distinguish between user=student (who uses what is served up) and user=teacher (who customises what is served up)

@@

R Commander should enter the discussion around here (even though it isn’t online)

P 5: Fig 2 shows a simple design that should be easy for casual users (as intro students really are) to get their heads around. Not all that keen on the use of “statistics” as a category heading in Panel 2 because it reinforces the ghettoizing of the word conflicting with a message that says everything you are doing here is statistics.

P 6 top: When you look at the actual R code generated it can be quite complex with its pipe structures etc which I would have thought does rather limit its introductory value.

**We appreciate the comment. In teaching R Workshops at Iowa State, we observed that among the majority of the introductory R students we encountered, the pipe structure was often preferred to using traditional nested function calls. The pipe structure is also the standard by which `ggvis` plots are created, which we are using as the backend to our graphical module. It would also be a relatively simple task for an instructor to modify or remove the piping structure in a modified version of `intRo` for use in a particular class.**

@@

P 6-7: discussion of particular modules should be in a “for example” vein. There is no need to justify what particular features are and are not included. (That would be in a teaching paper.)

**We have removed the in-depth discussion about the particular modules, and instead more vaguely described the functionality. We have also made sure to indicate that the full documentation is available online, which covers the supported functionality of every provided module.**

@@

P 7 top: Being able to choose only plot types that are “compatible” with the variables used does, however, preclude learning experiences of the form “this thing you want to do is not appropriate for this type of data.”

**We have also gone back and forth on this issue. The reasoning behind choosing only plot types compatible is that the user will see the different plot types available as they select different variables. So while the teaching moment could take place post-selection (by, for instance, producing an error if a plot type is used with incompatible variable types), we chose to have the teaching moment be a step earlier, where the student may see the list and wonder why the plot type they thought they wanted is not actually an option. Further, by doing a transformation of a variable, the plot types will automatically appear if they are compatible.**

@@

P 8: I think this descends to details too quickly. The big features are designing for extensibility, with conceptions about switching modules on and off and adding modules, and the interface as presented being assembled on the fly when the system starts up.

**We agree. We have restructured this portion of the paper to merely discuss the motivation behind an extensible system, with mentions of disabling and enabling modules. We have moved the details to the Understanding intRo section with the details on modules.**

@@

P 9-10 looks like appendix stuff to me.

Section 3 discusses downloading onto a local machine before customising the hosted online system which should be the focus (again conception and issues involved before any details).

**We have moved this section to after the discussion on customizing intRo. It now comes just before the conclusion.**

P 11 near bottom: “as an application on the user’s account” puzzled me -- needs to say shinyapps.io account

**Done**

Section 3.2 mainly belongs in up-front motivation. Table 1 is an implementation mechanism.

**We have merged remnants of Section 3.2 into Section 2 / motivation. We have also drastically cut down the content to focus this paper more on the extensible systems side, rather than the teaching side. As such, we have also removed Table 1.**

P 14 top: “currently unsupported” justifications are more relevant to a teaching paper than a software design paper.

**We have removed this comment.**

P 14 Last para Section 3: could go into motivational discussion.

**We have moved this paragraph to the motivational discussion.**

Section 4.1 is a big part of what I see as the real story but descends to details too quickly.

Most of Section 4.2 looks to me like more motivation, except the interpolate stuff which is an implementation mechanism

**We have merged most of the content from the former Section 4.2 (on reproducibility) with Section 2, and removed a discussion on interpolation.**

Section 4.3: more “main story”.

**We agree and have emphasized this section as one of the two primary pieces to the main story, the other being modularity.**

### 2016-05-26 jcgs-paper-2016/paper.Rmd
Continue response

Upon the release of RStudio's Shiny [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. This in turn led us to create an introductory statistics application which we call `intRo`, available at http://www.intro-stats.com. `intRo` offers a number of key advantages over traditional statistics software, including ease of use and an aim to foster student interest in coding. Additionally, `intRo` stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. These advantages will be discussed at length in this text. The paper is structured as follows: section \ref{what-is-intro} introduces the application, its features and its usability. Section \ref{using-intro} describes the process of creating a classroom instance of `intRo` tailored to the individual instructor. Section \ref{understanding-intro} provides technical details on the implementation of `intRo`. Finally, section \ref{conclusions-and-future-work} discusses some future possibilities and limitations of the software.

@@

In the development of `intRo`, we focused on aspects of the user interface (UI) and output that make it easy to pick up without extensive training. We used large, easy to click icons in the page header to help students find what they need more easily. We also made the functionality available the minimal necessary for an introductory statistics course. Figure \ref{fig:user_experience} illustrates the simple steps a student takes to generate a result in `intRo`. In this instance, a user clicks on the Graphical tab to create a mosaic plot. The user sees the plot, and elects to click the save button to store the plot (and its corresponding code) to the final compendium.

@@

We've also created a documentation website that is consistent with the interface of the application. The documentation is available at http://gammarama.github.io/intRo and hyperlinked within the application itself.

@@

On the front end, user interaction with `intRo` is split into bitesize chunks that we call modules. Modules are self-contained pieces of functionality which implement common statistical procedures that are often used in introductory statistics classes.

Within the data category there are two modules, sources and transformation. The data sources module allows the user to select or upload a dataset. Several datasets are pre-included, which can be selected via a drop-down menu. The user can check a box to upload a dataset, and then select the CSV file from her computer. Upon selection or successful upload of a dataset, a table view of the data will be visible on the right. From this module the user can select a subset of the data either by value or random selection. The dataset selected in the data module will be made available to all other modules of `intRo`. The data transformation module allows transformation of numeric variables to categorical and categorical variable to numeric. Additionally, numeric variable can be transformed using a power transformation. The transformed variables can be saved to the dataset for use elsewhere in `intRo`.

The graphical summaries module provides several common graphical displays of variables in the dataset. First, the user selects the plot desired, and either a single variable or two variables from the dataset, depending on plot type. The variable available to choose will only be those compatible with plot type selected. The plot visible on the right will automatically update when these values are selected. The numeric summaries module is another method of computing summaries of the data. The user is presented with a list of each variable from their dataset. The user can select one or more different variables, and will be presented with a table of some summary statistics. The summary statistics displayed vary depending on the type of the variables. For instance, for a numeric variable, a five number summary will be displayed. Character or factor variables will display counts of each unique value of that particular variable. Additionally, the user has the ability to display summaries by a grouping variable.

Within the set of statistics modules, there are three simple statistics tasks that an introductory statistics student will learn in her first semester. The contingency table module allows the user to compute a contingency table for two categorical variables. The results can either be displayed as counts, row percentages, column percentages, or total percentages. The simple linear regression module allows the user to compute a simple linear regression of a dependent variable (y) on an independent variable (x). The module will automatically filter any variables that are non-numeric so that the user may only select variables that are suitable for a simple linear regression. Results of the regression will be displayed on the right. A table of parameter estimates, a scatterplot of the data along with a line of best fit, The R-Squared value, and residual diagnostic plots will all be displayed. A user also has the ability to save residuals and predicted values to their dataset for further analysis. The T test module allows the user to perform either a one variable or two variable Student's T test. The user can select configuration options, including the direction of the test and the confidence level of the test. The results, including the t statistic, degrees of freedom, p-value, and a confidence interval, will be displayed in a text box in the results panel.

@@

# Using and Deploying `intRo`

>>>>

Upon the release of RStudio's Shiny [@shiny] it became easier for an R-based analysis to be converted to an interactive web application. This in turn led us to create an introductory statistics application which we call `intRo`, available at http://www.intro-stats.com. `intRo` offers a number of key advantages over traditional statistics software, including ease of use and an aim to foster student interest in coding. Additionally, `intRo` stands apart from new tools in that it is a supplement to an existing class, fully usable by a beginning statistics student. These advantages will be discussed at length in this text. The paper is structured as follows: section \ref{what-is-intro} introduces the application, its features and its usability, and provides motivations for the design decisions underlying it. Section \ref{understanding-intro} provides technical details on the design decisions of `intRo`, in particular, its modular and extensible structure. Section \ref{deploying-intro-instances} describes the process of deploying a classroom instance of `intRo` tailored to the individual instructor. Finally, section \ref{conclusions-and-future-work} discusses some future possibilities and limitations of the software.

@@

In the development of `intRo`, we focused on aspects of the user interface (UI) and output that make it easy to pick up without extensive training. We used large, easy to click icons in the page header to help students find what they need more easily. We also made the functionality available the minimal necessary for an introductory statistics course. Figure \ref{fig:user_experience} illustrates the simple steps a student takes to generate a result in `intRo`. In this instance, a user clicks on the Graphical tab to create a mosaic plot. The user sees the plot, and elects to click the save button to store the plot and its corresponding code to the final compendium.

@@

We've also created a documentation website that is consistent with the interface of the application. The documentation is available at http://gammarama.github.io/intRo and hyperlinked within the application itself. The documentation describes the specific functionality provided by the default `intRo` modules.

Statistics courses with a lab component can most benefit from the use of `intRo`. As students learn the aforementioned concepts, they can practice using the tools to perform the statistical tests and summaries. They simply upload a dataset and click the module appropriate to the section of the class. Once they choose the options desired by the instructor (variables, confidence levels, etc.), they can view the results and ultimately print the output for turning in. Early labs can walk the user through the limited number of button presses and clicks necessary, while later labs can more vaguely describe the intended output in hopes that the user can easily navigate the `intRo` interface to obtain the results.

@@

On the front end, user interaction with `intRo` is split into bitesize chunks that we call modules. Modules are self-contained pieces of functionality which implement common statistical procedures that are often used in introductory statistics classes. These modules form the core of `intRo` and are discussed at length in the next section.

@@

# Deploying `intRo` Instances
